Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-22 agujournaltemplate.tex
Intro rewrite done

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. The paraphyletic group consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}. Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial to understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Unfortunately, such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

@@

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and phytoplankton productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE
    Previous analyses of the phytoplankton community did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as 
    "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation" (AMO). 
    From this background, we aim to 1) investigate and relate changes in the depth distribution of biomass and community diversity over the entire time series, 2) check the structure and temporal changes of the phytoplankton community and 3) check whether in-situ upwelling metrics or climatic indices are better predicotrs of community change...
    % - microscopy data was not analyzed in detail, particularly towards time series of diversity.
    % Question: Does phytoplankton diversity follow the trends in biomass?
    We aim to investigate and relate changes in the depth distribution of biomass and community diversity over the entire time series.

    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE
    Previous analyses of the phytoplankton community did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as 
    "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation" (AMO).

    The impact of large-scale climatic processes on the observed shift in the community is still unclear. 
    %ENSO MEI v.2
    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.

    \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects in the tropical Atlantic \cite{enfield_tropical_1997}.

    % AMO??
    AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.

%% Gaps that We Adresse:
    % - More in-depth exploration of microscopy data & diversity time series
    % - Unidirectional trends driven by anthropogenic climate change or low-freq natural cycles, such as AMO?
        %Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
    % - Effect if ENSO on Phyotplankton community test, since previously not observed!
             %This could also be part of the rational of the study. Formulated as question, e.g. "it is still unclear the effects of regional climate phenomena such as ENSO on this shift" - Esteban
        % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."

    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression.

    Factors act at different scales \cite{wolkovich_temporal_2014}
    We aim to examine correlations and structures in time series data at monthly and annual time scales.

    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.

    - What are the open questions?? Drivers of community change.
    - Zooplankton/Sardines one important aspect, but data coverage is sparse.
    - Climate impacts, local upwelling system, etc.

    - Microscopy data in more detail (use entire depth and time)

    % FINAL PARAGRAPH LEADING TO METHODS
    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the relationships between bottom-up drivers, climate indices and the phytoplankton community. We focus on cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies. We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton. Our analysis emphasizes bottom-up drivers of community changes, as these data are available throughout the time series and at a monthly resolution. This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for comprehending the broader ecological changes in tropical marine ecosystems driven by shifts in climate.

>>>>

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. The paraphyletic group consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}. Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial to understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

@@

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and increased numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. This return in 2014 to conditions similar to those of the late 1990s coincided with a recovery in sardine catches observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}. 
    The impact of large-scale climatic processes on the observed shifts in the community is still unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important indicator of the climate system and its effects on the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
    The El Niño Southern Oscillation (ENSO) has a teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and did not test for correlations with the AMO index.

    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation."
    Previous analyses of phytoplankton community diversity did not include time series data beyond 2011, and so far there was no detailed investigation of changes in diversity based on the microscopy cell count data.
    From this background, we aim to 1) investigate changes in the depth distribution of phytoplankton biomass and community diversity over the entire time series, 2) relate the structure of the phytoplankton community data to the shifts observed in physical and biogeochemical variables, and 3) assess how decadal-scale climate osciallations as signified by the ENSO MEI v.2 and AMO index relate to changes in the phytoplankton community.
    Our analysis focuses exclusively on bottom-up drivers of community changes, as these data are available throughout the entire time series and at a monthly resolution. This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for comprehending the broader ecological changes in tropical marine ecosystems driven by shifts in climate.
